{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-07-31 submit/uist25-52/Source/paper.tex
Replace tikz for circled letters
TAPS revisions

\usepackage{xcolor,colortbl}
\usepackage{fontawesome}
\usepackage{enumitem}
\usepackage{fancyvrb}
\usepackage{tikz}
%\usepackage[breakable,theorems,skins]{tcolorbox}
%\usepackage{lettrine}
\usepackage{float}
\usepackage{hyperref}

@@

\newcommand*\circled[1]{\textnormal{\footnotesize\sffamily\bfseries\protect\tikz[baseline=(char.base)]{
  \node[shape=circle,fill=black,text=white,draw,inner sep=1pt] (char) {#1};}}}

% \DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
%   boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
%   colback=gray!20,colframe=gray!20,
%   enlarge bottom by=0pt,enlarge top by=0pt,
%   arc=0pt,outer arc=0pt]
% \lettrine[lraise=0.3]{\LARGE #1}{~}
% \small \textbf{#2.} #3
% \end{tcolorbox}
% }
\DeclareRobustCommand{\keyideabox}[3]
{\vspace{\dimexpr\baselineskip\relax} \noindent\colorbox{gray!20}{
\parbox{\dimexpr\columnwidth-\marginparsep+1pt\relax}
{\small {#1} \textbf{#2.} #3}
}}

@@

\begin{abstract}
\note{Added overview figure with examples showing relationships between Denicek \& co.}
User-centric programming research gave rise to a variety of compelling programming experiences,
including collaborative source code editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and system.

@@

A computational substrate defines the structures with which programs are constructed, how
the program state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
The choice of a substrate affects what programming experiences can be readily supported.
For example, object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory~\cite{steele-1993-lisp}.

@@

To illustrate merging, consider a case where we created a list of work items {\small\texttt{/todo}}.
In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
list in an extra \ident{<div>} element ($e_1$) and add a checkbox to each work item ($e_1'$):
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
e_2 &=& \srckvd{Append}(\textnormal{\small\texttt{/todo}}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
reflect the wrapping (2) and we need to create and additional \ident{Add} operation
that will add the checkbox to the new item (1). The result is a sequence with two edits $e_2', e_2''$
that target the newly wrapped list and add the checkbox to the added item:
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
  e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
e_2'  &=& \srckvd{Append}(\textnormal{\small\texttt{/todo/items}}, \#1, \#0, \srcid{<li>Do some work</li>})\\
e_2'' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/\#1}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we performed the merge operation in the other order, we would append $e_1, e_1'$ after
$e_2$ without a change. In this case, the result of applying the two edit sequences would be the
same.

@@

The key claim of this paper is that the Denicek computational substrate makes it easy to support
a range of compelling user experiences. The experiences can be implemented by composing the
primitive operations of the substrate, relying in particular on the operation for merging edits.
We demonstrate this claim with the web-based Webnicek system, which uses Denicek to support:
%
\begin{enumerate}[itemsep=0pt,label=(\roman*),topsep=3pt,leftmargin=20pt]
\item collaborative document editing (\S\ref{sec:impl-collab}),
\item programming by demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
\item incremental recomputation (\S\ref{sec:impl-eval} and \S\ref{sec:impl-incremental}),
\item schema change control (\S\ref{sec:impl-schema}),
\item end-user debugging via provenance tracking (\S\ref{sec:impl-provenance}),
\item concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
\end{enumerate}
%
We describe the programming experiences in isolation in the context of Webnicek in this section.
The data science notebook system Datnicek, discussed in \S\ref{sec:case}, provides
a more comprehensive case study by combining multiple user experiences together.

>>>>

\usepackage{xcolor,colortbl}
\usepackage{fontawesome}
\usepackage{enumitem}
\usepackage{fancyvrb}
% \usepackage{tikz}
%\usepackage[breakable,theorems,skins]{tcolorbox}
%\usepackage{lettrine}
\usepackage{float}
\usepackage{hyperref}

\usepackage{acmart-taps}

@@

% \newcommand*\circled[1]{\textnormal{\footnotesize\sffamily\bfseries\protect\tikz[baseline=(char.base)]{
%   \node[shape=circle,fill=black,text=white,draw,inner sep=1pt] (char) {#1};}}}

\newcommand*\circled[1]
{{\Large \textcircled{\footnotesize {#1}}}}

% \DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
%   boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
%   colback=gray!20,colframe=gray!20,
%   enlarge bottom by=0pt,enlarge top by=0pt,
%   arc=0pt,outer arc=0pt]
% \lettrine[lraise=0.3]{\LARGE #1}{~}
% \small \textbf{#2.} #3
% \end{tcolorbox}
% }
\DeclareRobustCommand{\keyideabox}[3]
{\vspace{\dimexpr\baselineskip\relax} \colorbox{gray!20}{
\parbox{\dimexpr\columnwidth-\marginparsep+1pt\relax}
{\small {#1} \textbf{#2.} #3}
}}

@@

\begin{abstract}
\note{Added overview figure with examples showing relationships between Denicek \& co.}
User-centric programming research gave rise to a variety of compelling programming experiences,
including collaborative source code editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and system.\enlargethispage{12pt}

@@

A computational\enlargethispage{12pt} substrate defines the structures with which programs are constructed, how
the program state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
The choice of a substrate affects what programming experiences can be readily supported.
For example, object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory~\cite{steele-1993-lisp}.

@@

To illustrate merging, consider a case where we created a list of work items {\small\texttt{/todo}}.
In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
list in an extra \ident{<div>} element ($e_1$) and add a checkbox to each work item ($e_1'$):
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
e_2 &=& \srckvd{Append}(\textnormal{\small\texttt{/todo}}, \#1, \#0, \srcid{<li>Do some work</li>})\\
e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
reflect the wrapping (2) and we need to create and additional \ident{Add} operation
that will add the checkbox to the new item (1). The result is a sequence with two edits $e_2', e_2''$
that target the newly wrapped list and add the checkbox to the added item:
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
  e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
e_2'  &=& \srckvd{Append}(\textnormal{\small\texttt{/todo/items}}, \#1, \#0, \srcid{<li>Do some work</li>})\\
e_2'' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/\#1}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we performed the merge operation in the other order, we would append $e_1, e_1'$ after
$e_2$ without a change. In this case, the result of applying the two edit sequences would be the
same.

@@

The key claim of this paper is that the Denicek computational substrate makes it easy to support
a range of compelling user experiences. The experiences can be implemented by composing the
primitive operations of the substrate, relying in particular on the operation for merging edits.
We demonstrate this claim with the web-based Webnicek system, which uses Denicek to support:
%
\begin{enumerate}%[itemsep=0pt,label=(\roman*),topsep=3pt,leftmargin=20pt]
\item[(i)] collaborative document editing (\S\ref{sec:impl-collab}),
\item[(ii)] programming by demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
\item[(iii)] incremental recomputation (\S\ref{sec:impl-eval} and \S\ref{sec:impl-incremental}),
\item[(iv)] schema change control (\S\ref{sec:impl-schema}),
\item[(v)] end-user debugging via provenance tracking (\S\ref{sec:impl-provenance}),
\item[(vi)] concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
\end{enumerate}
%
We describe the programming experiences in isolation in the context of Webnicek in this section.
The data science notebook system Datnicek, discussed in \S\ref{sec:case}, provides
a more comprehensive case study by combining multiple user experiences together.
